\section{Introduction}
\label{sec:introduction}

Session types
\cite{DBLP:conf/concur/Honda93,DBLP:conf/parle/TakeuchiHK94,DBLP:conf/esop/HondaVK98}
have become a quasi-standard for defining stateful communication
protocols and reasoning about them. A (binary) session
type specifies a protocol as a set of sequences of typed messages going forth
and back between two participants. Many interesting protocols are
recursive, which requires nontrivial reasoning about equivalence and
subtyping of session types. For context-free session types
\cite{DBLP:conf/icfp/ThiemannV16}, equivalence is expensive to decide
and subtyping is undecidable. This paper studies algebraic representations of session types that
admit linear time equivalence and subtyping while retaining most of the
expressivity of context-free session types.

Traditionally, protocols are modeled using equirecursive session
types like \lstinline{IntListS} (session type for sending finitely many integers and then
close) and \lstinline{IntStreamS} (send infinitely many integers) that
are assigned to one end of a communication channel:
\begin{lstlisting}
  type IntListS   = mualpha. oplus{Cons: !Int.alpha, Nil: EndT} -- finite sequences
  type IntStreamS = mualpha. !Int.alpha                    -- infinite sequence
\end{lstlisting}
In session types, the $\oplus\{l_i:S_i\}$ constructor stands for
selection: send a label $l_i$, like
\lstinline|Cons| or \lstinline|Nil|, and continue according to the
chosen branch $S_i$; the $!T.S$ constructor stands for sending a value of
type $T$ and continuing the protocol according to $S$. The other end
of the channel has the \emph{dual type} where sending and receiving ($?T.S$)
is flipped, as well as selection and branching indicated by
$\&\{l_i:S_i\}$: if we receive a label $l_i$, then continue with $S_i$.

Using recursive types directly is somewhat tedious as equivalence of
equirecursive types\footnote{An equirecursive type is considered
  \emph{equal} to its unfoldings.} amounts to comparing infinite regular trees or
reasoning up to unfolding of the recursive types. While equivalence is
decidable in $O(n^2)$ for traditional, tail-recursive session types
(where $n$ corresponds to the size of the type), 
the difficulty increases when considering context-free session types
\cite{DBLP:conf/icfp/ThiemannV16,DBLP:conf/esop/Padovani17,DBLP:journals/toplas/Padovani19,DBLP:journals/corr/abs-2106-06658,
  DBLP:conf/esop/DasDMP21} where the first equivalence test was doubly
exponential\cite{DBLP:conf/tacas/AlmeidaMV20}, though recent work
indicates it might be polynomially decidable in
$O(n^{12})$ \cite{DBLP:journals/corr/abs-2407-04063}.

These results expose a tension between expressivity and
tractability: richer session type languages lead to expensive
equivalence and undecidable subtyping. Our goal is to identify a
representation that retains expressivity close to context-free session
types while admitting efficient equivalence and subtyping.

Prior work proposed parameterized algebraic
protocols \cite{DBLP:journals/pacmpl/MordidoS0V23}, where recursive
protocols are defined analogously to algebraic datatypes (ADTs).
A protocol is defined by a nominal type whose
constructors describe the possible next actions. 
In this framework, we
can define \lstinline{IntListP}, the algebraic protocol analogous
to \lstinline{IntListS}, and \lstinline{IntStreamP}, analogous to
\lstinline{IntStreamS}, as follows:
\begin{lstlisting}
  protocol IntListP   = Nil | Cons Int IntListP
  protocol IntStreamP = Next Int IntStreamP
\end{lstlisting}
As with ADTs, the combination of recursive types with sum-of-product
types is ``shrink wrapped'' into a nominal type. In consequence,
deciding equivalence of protocol types takes time linear in the size
of the type, regardless of whether the underlying protocol is
tail-recursive or context free.

The prior work gave an informal outline of an embedding of
algebraic protocols into context-free session types, but it did not offer
a comparison with traditional, tail-recursive session types. This paper fills this gap.

\subsection{Traditional Session Types to Algebraic Protocols}
\label{sec:trad-sess-types}

We now give an outline of a translation from traditional session types
(TST) to algebraic session types (AST).
The translation replaces equirecursive types by explicit
systems of recursive equations.
The main difficulty in translating a program with  TST to AST lies in managing the unfolding of
equirecursive types. That is, at every point in a TST typing
derivation we may silently apply a type conversion that unrolls a
recursive session type as in $\mu\alpha.S \approx S[\mu\alpha.S /
\alpha]$. Our translation to AST proceeds as follows.
First, we translate \emph{all} types in a TST program into a system of recursive
type equations where all left hand sides are type variables and
equations are flat, i.e., they consist of a type constructor applied
to type variables as in
\lstinline{alpha->beta} or \lstinline{?alpha.beta}. All type variables
occurring on right hand sides must be defined by a single
equation. Defined type variables on the left hand sides may be used
zero or more times on the right hand sides. Here are two equation
systems (one line each)
corresponding to our running examples.
\begin{lstlisting}
IntStreamS  alpha = !beta.alpha                   beta = Int
IntListS    alpha = (+){Cons: beta, Nil: delta}   beta = !gamma.alpha    gamma = Int   delta = End
\end{lstlisting}
Such an equation system can be considered as a
labeled transition system (LTS). Each type variable corresponds to a
state and each equation induces outgoing labeled transitions. For
example, in the LTS corresponding to \lstinline{IntListS}, there are
labeled transitions $\alpha \stackrel{+Cons}{\longrightarrow} \beta$ and $\alpha \stackrel{+Nil}
{\longrightarrow} \delta$ and further transitions for $\beta$,
$\gamma$, and $\delta$. Generally, the labeled transitions indicate
the top-level type constructor and the particular step taken. In a second step, we minimize this
LTS. Minimization corresponds to equating bisimilar type
variables, where bisimilarity corresponds to type equivalence of
equirecursive types. Thus, after minimization two types are bisimilar
if and only if they are represented by the \emph{same} type
variable. Consequently, we can eliminate the conversion rule from
typing as two types are now equivalent iff they are represented by the
same type variable.

The final step of the translation reads algebraic protocols from
the equations. To this end, we generate a decision protocol for each
type variable with a select type or branch type (like $\alpha$ for
\lstinline|IntListS|) and use predefined protocols for sequence, empty
protocol, and repetition to translate sending or receiving session
types. For \lstinline|IntListS| and \lstinline|IntStreamS|, our
translation yields
the algebraic protocols \lstinline{IntListSP} and
\lstinline{IntStreamSP}, shown after the definitions of the auxiliary protocols.
\begin{lstlisting}
  protocol Seq a b = Seq a b              -- first a then b
  protocol Eps = Eps                      -- empty protocol
  protocol Repeat a = Repeat a (Repeat a) -- infinitely often a
  protocol List a = Cons a (List a) | Nil -- select/branch: Cons or Nil

  IntListSP   = List (Seq Int Eps)
  IntStreamSP = Repeat (Seq Int Eps)
\end{lstlisting}
Here the \lstinline+List+ protocol is the only custom protocol
generated by the translation. The other protocols \lstinline+Seq+,
\lstinline+Eps+, and \lstinline+Repeat+ are predefined.

The resulting algebraic protocols are bisimilar to the
original session types when both are interpreted as labeled
transition systems. We consider this outcome as a statement of
\emph{interoperability} between the original TST program and its AST
translation.  Interoperability results
from an innocuous change that was hinted at, but not executed in the
prior work \cite{DBLP:journals/pacmpl/MordidoS0V23}: single
constructor protocols do not send or receive their constructor. That
is, the protocols \lstinline|Seq|, \lstinline|Eps|, and
\lstinline|Repeat| are tag-free as they do not transmit a tag.
In our example, it even holds that all three representations of the
types are bisimilar.
\begin{lstlisting}
  IntListS   ~ IntListSP   ~ IntListP
  IntStreamS ~ IntStreamSP ~ IntStreamP
\end{lstlisting}

\subsection{Subtyping for Algebraic Protocols}
\label{sec:subtyp-algebr-prot}


We define a subtyping relation on parameterized algebraic
protocols. This subtyping relation is inspired by prior work on
constructor subtyping \cite{DBLP:conf/esop/BartheF99}, where a
% restrictable variants \cite{DBLP:conf/ecoop/MadsenSL23}
subtype of an algebraic data type is defined by a non-empty subset of its
constructors. In our framework, the protocol \lstinline{IntListP} comes in
three variants: \lstinline{IntListP[Nil]}, \lstinline{IntListP[Cons]},
and \lstinline{IntListP[Nil,Cons]} ($=$ \lstinline{IntListP}). They roughly correspond to the
session types \lstinline|oplus{Nil: EndT}|,
\lstinline|mualpha.oplus{Cons: !Int.alpha}|, and
\lstinline{IntListS}. To define the subtyping relation on these
variants, we have to give the protocol a \emph{direction}: the type
\lstinline{!IntListP} interprets the protocol in the forward direction
(corresponding exactly to \lstinline{IntListS}), whereas
\lstinline{?IntListP} interprets the protocol in the reverse direction
(corresponding exactly to the dual of \lstinline{IntListS}). Taking
directionality into account, we obtain the following subtyping judgments:
\begin{lstlisting}
!IntListP <: !IntListP[Nil]                ?IntListP[Nil]  <: ?IntListP
!IntListP <: !IntListP[Cons]               ?IntListP[Cons] <: ?IntListP
\end{lstlisting}
The full subtyping relation corresponds to the reflexive transitive
closure of the relation defined by these judgments.

The subtyping relation resulting from constructor subtyping on algebraic protocols
is still decidable in linear time in the size of the type if comparing
two sets of constructors takes constant 
time\footnote{This assumption is reasonable because the number of
  constructors is usually bounded by a small constant and a set of
  constructors can be represented by a bit set.}. In comparison,
subtyping for tail-recursive session types is decidable in $O(n^2)$ and
subtyping for context-free session types is undecidable
\cite{DBLP:journals/toplas/Padovani19}. That is, 
algebraic protocols with constructor subtyping carve out a subset of
context-free session types with an efficiently decidable subtyping relation.

\subsection{Traditional Session Types with Subtyping to Algebraic Protocols}
\label{sec:trad-sess-types-1}


The constructor subtyping relation is key in extending the mapping from traditional,
tail-recursive session types to algebraic protocols to a setting with
subtyping. The first step in constructing this mapping is just like in
\cref{sec:trad-sess-types}: we represent all types in a program by a system of type
equations. The second step is also the same: we minimize the
equations so that each type variable represents an equivalence class
of bisimilar types.

In our model, 
a type is a pair $\langle\alpha, \mathcal{E}\rangle$ where
$\mathcal{E}$ is a system of type equations and $\alpha$ is the left
hand side of an equation in $\mathcal{E}$. To stay in this framework,
we define subsumption on type equations: Subsumption keeps the
variable $\alpha$; it only adapts $\mathcal{E} <: \mathcal{E}'$ where
the standard subsumption may be applied to any suitable equation. For instance,
here is the subsumption from \lstinline{IntListS} to
\lstinline{IntListS[Nil]} which eliminates the constructor
\lstinline{Cons}: 
\begin{lstlisting}
IntListS         alpha = (+){Cons: beta, Nil: delta}   beta = !gamma.alpha    gamma = Int   delta = End  
IntListS[Nil]    alpha = (+){Nil: delta}            beta = !gamma.alpha    gamma = Int   delta = End
\end{lstlisting}
The domains of $\mathcal{E}$ and $\mathcal{E}'$ coincide, i.e., all
left hand sides stay the same and we do not drop 
unreachable type variables like $\beta$ and $\gamma$. Getting from the
right hand side of $\alpha$ in the first line to the second line
consists in applying the standard subsumption rule for selection in
functional session types \cite{DBLP:journals/jfp/GayV10}. In the
general case, we may apply subsumption to each equation individually,
taking care whether the left hand side variable occurs in covariant
(like in our example) or in contravariant position.

\subsection*{Contributions}

This paper develops algebraic protocols with constructor
subtyping and shows that they provide an efficient and expressive
framework for typing session-based programs.
Algebraic protocols identify a tractable fragment of context-free
session types with linear time algorithms for equivalence and subtyping.


\begin{enumerate}

\item
We define a subtyping relation for algebraic protocol types based on
constructor subtyping. Subtyping is obtained by selecting non-empty
subsets of protocol constructors. The symmetric kernel of this
relation coincides with the equivalence relation defined in prior work
\cite{DBLP:journals/pacmpl/MordidoS0V23}.  Subtyping and type
equivalence are decidable in linear time.

\item
We define the calculus \algstsub, a core language based on linear
System~F with non-linear recursion and session types expressed as
algebraic protocols. It extends \algst from prior work with constructor
subtyping. Typing rules are presented in bidirectional
form and admit a direct algorithmic interpretation.

\item
We establish type soundness for \algstsub via preservation and
progress.

\item
We define a translation from traditional tail-recursive session
types to \algst.  The translation preserves typing and establishes
a reduction correspondence.

\item
We extend this translation to session types with subtyping under a
restriction on the representation of session types as recursive
equation systems.

\item
We provide an implementation of the system as a publicly available
artifact, including practical extensions described in
\cref{sec:implementation}.

\end{enumerate}



\subsection*{Overview}
\label{sec:overview}

TODO

% \Cref{sec:algebr-sess-types} contains an informal introduction of algebraic
% protocols with examples; we illustrate the gain in
% expressivity and modularity with respect to other work in
% \cref{sec:param-prot}. \Cref{sec:types} discusses the type structure
% of \algst and establishes key results for efficient type checking: correctness of type normalization and
% worst case complexity of type equivalence. \Cref{sec:processes} defines
% statics and dynamics of expressions 
% and processes % . \Cref{sec:metatheory}
% followed by a presentation of the metatheoretical results.
% % \Cref{sec:cfst-vs-algst} gives an informal outline of
% % an embedding of context-free session types to \algst.
% \Cref{sec:implementation} discusses the implementation; since all rules are
% algorithmic, their appropriate reading leads directly to a type checker and an interpreter.
% \Cref{sec:related,sec:conclusion} discuss related work and conclude.
% %
% Proofs, auxiliary results, further examples, an informal outline of an embedding
% of context-free session types to \algst, and further discussion are provided as supplementary material. The
% implementation is publicly available \cite{artifact}.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
