% \section{Metatheory}
% \label{sec:metatheory}

% This section introduces the main results of our language: preservation and
% progress possibly ending in a deadlock.

\input{fig-proc-formation}

We complete section with the main results of our language.
%
Towards this end we need to talk about context and process typing.
%
Type formation (\cref{fig:type-formation}) is lifted pointwise to contexts in 
judgement $\isCtx{\Gamma}{\kind}$.
%
% , defined by the rules below. Notice that the
% last premise in rule Ctx-Empty checks the invariant whereby contexts contain
% only normalized terms.
% %
% \begin{mathpar}
%      \infrule{Ctx-Empty}
%      {}
%      {\isCtx{\Empty}{\kind}}
%  
%      \infrule{Ctx-Ext}
%      {\isCtx{\Gamma}{\kind}\\ \typeAgainst{T}{\kind} \\ \typec T = \Normal[+]{T}}
%      {\isCtx{\Gamma, \ebind{x}{T}}{\kind}}
% \end{mathpar}
%
The rules for typing processes are in
\cref{fig:proc-formation}. The rule for threads requires the expression
$\termc e$ to consume all resources given in context $\Gamma$. Rule P-Par splits
the context in two parts and gives each to one of the parallel processes. The
rule is not algorithmic but we plan to type check expressions only; processes
are a runtime artefact. Rule P-New guesses a type $\typec T$ for channel end
$\termc x$, then channel end $\termc y$ gets the type $\TDual T$, both in normal
form. Again, this rule is not algorithmic and need not be.
%
The labelled transitions convey behaviour in their labels. Because we use an
algorithmic typing system with leftovers, we require a labelled transition
system for typing contexts to account for the impact of the behaviour of labels
in the contexts. The statement of preservation relies on the labelled transition
system for contexts, defined in \cref{sec:proofs} (\cref{fig:lts-ctx}) 
\hl{[\text{MW}]:} and subtyping lifted to contexts where elements in the context are compared pairwise.

% \hl{[\text{MW}]:} Because our type system incorporates subtyping and pointwise joins for contexts, 
% type preservation for synthesis no longer guarantees strict type equivalence. 
% Instead, a well-typed expression that takes a step will synthesize a type $\typec{T'}$ 
% that is a subtype of the original type $\typec{T}$.

\begin{theorem}[Preservation]\
  \label{thm:preservation}
  \begin{enumerate}
  \item\label{it:pres-synth} If $\trans{e_1}{\lambda}{e_2}$ and
    $\expSynth{\Gamma_1,\Gamma_2,\Gamma_3}{e_1}{T}{\Gamma_3}$ and
    $\transCtx{\Gamma_2}{\lambda}{\Gamma_2'}$ and 
    $\isCtx{\Gamma_1,\Gamma_2,\Gamma_3}{\kindt}$, then
    % MW 
    % Note: I think we need the \Gamma_3' \subt \Gamma_3.
    % Also, wouldnt it have need to be \Gamma_3' \equiv_\alpha \Gamma_3 before?
    $\expSynth{\Gamma_1,\Gamma_2',\Gamma_3}{e_2}{T'}{\hl{\Gamma_3'}}$ and
    $\hl{\typec{T'} \subt \typec{T}}$ and $\hl{\Gamma_3' \subt \Gamma_3}$.
  \item\label{it:pres-against} If $\trans{e_1}{\lambda}{e_2}$ and
    $\expAgainst{\Gamma_1,\Gamma_2,\Gamma_3}{e_1}{T}{\Gamma_3}$ and
    $\transCtx{\Gamma_2}{\lambda}{\Gamma_2'}$ and 
    $\isCtx{\Gamma_1,\Gamma_2,\Gamma_3}{\kindt}$, then
    $\expAgainst{\Gamma_1,\Gamma_2',\Gamma_3}{e_2}{T}{\Gamma_3}$.
  \item\label{it:pres-proc} If $\trans{p_1}{\pi}{p_2}$ and
    $\isProc[\Gamma_1,\Gamma_2]{p_1}$ and $\transCtx{\Gamma_2}{\pi}{\Gamma_2'}$
    and $\isCtx[\Empty]{\Gamma_1,\Gamma_2}{\kindt}$,
    then $\isProc[\Gamma_1,\Gamma'_2]{p_2}$.
  \end{enumerate}
\end{theorem}

%\paragraph{Progress}

To the below set of labels we call the \emph{progress set} for $\termc x,\termc y$:
%
\begin{equation*}
\{
% processes
\tau,\,
(\parl{\receivel xv}{\sendl yv}),\,
(\parl{\receivel xC}{\sendl yC}),\,
(\parl{\closel x}{\openl y}),\,
(\parl{\receivel xa}{\scopel abya}),\,
(\parl{\receivel xb}{\scopel abyb})
\}  
\end{equation*}

\emph{Completed processes} are of the form
$\termc{\expp\unitk \PAR \dots \PAR \expp\unitk}$.
%
A process is a \emph{deadlock} if, for all its subterms of the form $\newp xyp$
and all transitions $\trans{p}{\pi}{p'}$, we have $\pi$ not in the progress set
for channel $\termc x,\termc y$.
%
To identify a deadlocked process we find all its $\nu$ subterms, and for each
subterm we find all its transitions. Then, if all transitions fall outside the
progress set, the process is deadlocked.
%
When the kind context $\Delta$ is apparent from the context, we write
$\Gamma^{\kinds}$ to denote a typing context composed solely of session types,
that is, such that $\ebind xT\in\Gamma$ implies $\typeAgainst T \kinds$.

\begin{theorem}[Progress possibly leading to deadlock]\
  \label{thm:progress}
  \begin{enumerate}
  \item\label{item:prog-exps} If
    $\expSynth{\Gamma_1^{\kinds}}{e_1}{T}{\Gamma_2}$, then either $\termc{e_1}$
    is a value or $\trans{e_1}{\lambda}{e_2}$.
  \item  If $\expAgainst{\Gamma_1^{\kinds}}{e_1}{T}{\Gamma_2}$, then either $\termc{e_1}$
    is a value or $\trans{e_1}{\lambda}{e_2}$.
  \item\label{item:prog-procs} If $\isProc[\Gamma^\kinds]{p_1}$, then
    $\termc{p_1}$ is completed, $\trans{p_1}{\pi}{p_2}$, or $\termc{p_1}$ is
    a deadlock.
  \end{enumerate}
\end{theorem}
%


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
