\section{Algebraic Protocols, Session Types, and Subtyping}
\label{sec:algebr-sess-types}

This section starts with a brief introduction to algebraic protocols
driven by examples. Next, it motivates constructor subtyping via
compatibility for evolving interfaces. The remaining sections discuss
the translation from TST (traditional, tail-recursive session types)
to algebraic session types with examples, first without subtyping and
finally with subtyping. We use a
syntax inspired by Haskell, as implemented in our artifact. Examples make
liberal use of practical extensions of the formal system (cf.\ \cref{sec:implementation}).

\subsection{Algebraic Protocols}
\label{sec:algebr-datatyp-as}

% Suppose we want to transmit abstract syntax trees for a language with integers
% and addition. Their type might be defined by an algebraic datatype definition as
% in Haskell or ML.
% \begin{lstlisting}
% data Ast = Con Int | Add Ast Ast
% \end{lstlisting}
% This declaration defines the type \lstinline+Ast+ along with the types of the constructor
% functions, which can be used to construct \lstinline+Ast+ values as well as for pattern matching
% on them:
% \begin{lstlisting}
%   Con :        Int -> Ast
%   Add : Ast -> Ast -> Ast
% \end{lstlisting}
% The type \lstinline+Ast+ is recursive and its \lstinline+Add+
% constructor takes two \lstinline+Ast+ values as arguments.
% To define a protocol for transmitting values of type \lstinline+Ast+, we merely change the keyword
% \lstinline+data+ to \lstinline+protocol+ as in
% \begin{lstlisting}
% protocol AstP = ConP Int | AddP AstP AstP
% \end{lstlisting}
% This declaration defines the protocol type \lstinline+AstP+ along with tags \lstinline+ConP+
% and \lstinline+AddP+, which serve to introduce and eliminate branching in the protocol.
% % Tags are encoded by small numbers for transmission.
% The resulting \lstinline+AstP+ protocol can be used in
% two directions, \lstinline+!AstP+ for sending and \lstinline+?AstP+ for receiving
% data.\footnote{The implementation allows us to reuse a
%   \lstinline+data+ declaration as a \lstinline+protocol+ declaration,
%   thus overloading the constructors.} 
% Protocol types only describe behavior, they do not classify run-time values. They inhabit a
% dedicated kind \lstinline+KP+.

% We illustrate the \lstinline+AstP+ protocol with a function to send an
% \lstinline+Ast+ value. We write functions
% in \emph{channel passing style}, which takes a channel as its argument
% and returns it after processing some prefix of the channel's session
% type. The function's type is universally quantified over the continuation
% type \lstinline+s+ of the unexplored part of the channel.
% % This quantification must be introduced explicitly using
% % \lstinline+forall(s:S)+ where \lstinline+KS+ is the kind of session
% % types.
% The argument
% type adds some prefix to \lstinline+s+ that describes the part of the
% session performed by the function. In our example the prefix is \lstinline+!AstP+.
% \begin{lstlisting}
% sendAst : Ast -> forall(s:S). !AstP.s -> s
% sendAst t [s] c = case t of {
%     Con x   -> select ConP [s] c |> sendInt [s] x,
%     Add l r -> select AddP [s] c |> sendAst l [!AstP.s] |> sendAst r [s] }
% \end{lstlisting}

% We use the \lstinline+|>+ operator for reverse function
% application. Defined by \lstinline+x |> f = f x+, the operator associates
% to the left and features low priority, so that \lstinline+x |> f |> g+ means \lstinline+g (f x)+.
% %
% The square brackets indicate type abstraction (on the left) and type
% application (in expressions).
% % By abstracting over a type variable \lstinline+s+ of kind \lstinline+S+
% % (introduced by \lstinline+forall(s:S)+), we require \lstinline+s+
% % to be later replaced by a session type.
% %
% The predefined \lstinline+sendInt+ operation has type
% \lstinline+forall(s:S). Int -> !Int.s -> s+.
% %
% The use of pattern matching with  \lstinline+case+ on algebraic
% datatypes is standard.
% As usual in session type systems, channel values of session type are linear
% and the \lstinline+select+ operation makes an internal choice between alternatives. Selecting a protocol
% constructor sends an encoding of the corresponding tag and ``pushes'' the transmission of the
% fields on the channel type. That is:
% \begin{lstlisting}
%   select ConP [s] : !AstP.s -> !Int.s
%   select AddP [s] : !AstP.s -> !AstP.!AstP.s
% \end{lstlisting}
% A function to process the receiver end of this protocol must take an argument of type \lstinline+?AstP.s+.
% \begin{lstlisting}
% recvAst : forall(s:S). ?AstP.s -> (Ast, s)
% recvAst [s] c = match c with {
%     ConP c -> let (x, c)  = receiveInt [s] c    in (Con x, c),
%     AddP c -> let (tl, c) = recvAst [?AstP.s] c in
%               let (tr, c) = recvAst [s] c       in (Add tl tr, c) }
% \end{lstlisting}
% The \lstinline+match+ on a session-typed channel receives
% a tag and branches according to it. Branching also ``pushes'' transmission of the fields on the
% channel type analogous to the \lstinline+select+ operation:
% \begin{lstlisting}
% c : ?AstP.s |- match c with { ConP c -> ...   -- c : ?Int.s
%                               AddP c -> ... } -- c : ?AstP.?AstP.s
% \end{lstlisting}
% Left of the turnstile $\vdash$, we write the typing of
% \lstinline+c+ before the \matchk. In the body of the \matchk, we write
% the typing of \lstinline+c+ after matching against the selectors \lstinline+ConP+ and \lstinline+AddP+,
% respectively.
% The \lstinline+match c+ (with the type on the left) consumes the linear
% binding for channel \lstinline+c+. The pattern \lstinline+ConP c+
% reintroduces variable \lstinline+c+, binds it with the channel
% after processing the selection and gives it an updated type (on the right) for the body of the
% corresponding branch. The consideration for pattern \lstinline+AddP c+
% is analogous. The choice of the branch is external and depends
% on the received selector tag.

% The protocol in this subsection goes beyond the reach of most session type systems because
% they only support tail recursive session types.
% Thus, functions like \lstinline+sendAst+ and \lstinline+recvAst+ have
% no meaningful type in traditional
% session type systems \cite{DBLP:journals/jfp/GayV10,lindley17:_light_funct_session_types}.
% However, they can be written using context-free or nested session
% types \cite{DBLP:journals/corr/abs-2106-06658,DBLP:conf/icfp/ThiemannV16,DBLP:conf/esop/DasDMP21}. We
% discuss the differences in \cref{sec:related}.

% \subsection{Simple protocols with polarities}
% \label{sec:simple-protocols-add}

We start with a brief recap of algebraic protocols as presented by \citet{DBLP:journals/pacmpl/MordidoS0V23}.
As a running example, we choose the arithmetic server, which is a
standard example from the session type literature (cf.\ 
\cite{DBLP:conf/esop/GayH99}). The arithmetic server
accepts two commands, \lstinline+Neg+ and \lstinline+Quit+. When it
receives the \lstinline+Neg+ command, it waits for an integer
argument, responds with an integer (the result) and loops back to
accept the next command. When it receives \lstinline+Quit+, it terminates
the connection.

In the session type literature, the type of the arithmetic server is a
recursive type like this one.
\begin{lstlisting}
  ArithS = mualpha.&{Neg:?Int.!Int.alpha, Quit:end}
\end{lstlisting}

Written as an algebraic protocol, we obtain the definition of \lstinline+ArithP+.
\begin{lstlisting}
  protocol ArithP = Neg Int -Int | Quit
\end{lstlisting}
Unlike the session type \lstinline+ArithS+, \lstinline+ArithP+ is a
protocol. Here is
the crucial difference: 
The session type \lstinline+ArithS+ already fixes the direction of communication, as the
operators \lstinline+&+ and \lstinline+?+ indicate receiving. The
protocol \lstinline+ArithP+ only fixes the structure of the
communication, i.e., \lstinline+Neg+ and then two \lstinline+Int+
transmissions in the same direction while the \lstinline+-Int+ indicates a
transmission in the opposite direction. We materialize the direction
of a protocol by using it in a session type as in
\lstinline+?ArithP.End+. This algebraic session type corresponds exactly to
\lstinline+ArithS+. 

A client implementation would use the select operator as usual, but
with the following types dictated by the theory of algebraic
protocols. We generally write type arguments in square brackets.
\begin{lstlisting}
  select Quit [s] : !ArithP.s -> s
  select Neg [s]  : !ArithP.s -> !Int.?Int.!ArithP.s
\end{lstlisting}
Branching operates analogously on types in the assumption:
\begin{lstlisting}
c : ?ArithP.s |- match c with { Quit c -> ...  -- c : s
                                Neg c -> ... } -- c : ?Int.!Int.?ArithP.s
\end{lstlisting}

The implementation of the server is straightforward.
\begin{lstlisting}
serveArith : ?ArithP.end -> unit
serveArith c = match c with {
    Quit c -> close c,
    Neg c  -> let (x, c) = receiveInt [!Int.?ArithP.end] c in
              let c = sendInt [?ArithP.end] (0-x) c in
              serveArith c }
\end{lstlisting}
The type arguments to \lstinline+receiveInt+ and \lstinline+sendInt+
are the types of the respective continuation channels. They specify
the type of the variable \lstinline+c+ that is bound on the left side.

\begin{lstlisting}
  arithClient : !ArithP.end -> Int -> Int
  arithClient c x =
    let c = select Neg [end] c in
    let c = sendInt [?Int.!ArithP.end] x c in
    let (r, c) = receiveInt [!ArithP.end] c in
    let c = select Quit [end] c in
    let _ = close c in
    r
\end{lstlisting}

The main feature of algebraic protocols that we did not touch, yet, is
parametric protocols. That means that protocols can be parametric over
other protocols. This feature enables the definition of protocols like
\lstinline+Seq+ from the introduction. We introduce it in XXX along
with the discussion of the translation.

\subsection{Constructor Subtyping for Algebraic Protocols}
\label{sec:constr-subtyp-algebr}

Imagine an extension of the arithmetic server with a new operation,
say, addition indicated by the command \lstinline+Add+. The
traditional session type changes to include the new alternative.
\begin{lstlisting}
  ArithS' = mualpha.&{Add:?Int.?Int.!Int.alpha,Neg:?Int.!Int.alpha, Quit:end}
\end{lstlisting}
The standard subtyping for session types gives
\lstinline+ArithS <: ArithS'+ because a channel of type
\lstinline+ArithS+ can be processed by code that expects the supertype
\lstinline+ArithS'+: The code has an additional branch that is never
exercised.

With constructor subtyping for algebraic protocols, we obtain exactly
the same situation. The extended protocol type becomes
\begin{lstlisting}
  protocol ArithP = Add Int Int -Int | Neg Int -Int | Quit
\end{lstlisting}
Now,
\lstinline+ArithP[Neg,Quit]+ is a subtype of \lstinline+ArithP+ which
corresponds exactly to the previous version. 


\subsection{Parameterized protocols and modularity}
\label{sec:param-prot}

Algebraic protocols can have protocol parameters,  just like algebraic data
types have type parameters. This facility provides a powerful and versatile approach for declaring
modular protocols.
As a first example, we declare a protocol that repeatedly runs
another protocol:
\begin{lstlisting}
protocol Stream a = Next a (Stream a)
\end{lstlisting}
Using this protocol we define a process that sends infinitely many ones on a channel:
\begin{lstlisting}
ones : !Stream Int.EndT -> Unit
ones c = select Next [Int,EndT] c |> sendInt [!Stream Int.EndT] 1 |> ones
\end{lstlisting}
When using the \lstinline+Stream+ protocol, the continuation session
type \lstinline+EndT+ does not matter because the protocol
never reaches the continuation. For a parameterized protocol,
\lstinline+select+ takes type parameters corresponding to the protocol
parameters (\lstinline+Int+) as well as the continuation session (\lstinline+EndT+).
Algebraic protocols sidestep issues with contractiveness of
recursive definitions and thus guardedness of the protocols: we always
have to make explicit progress by selecting the \lstinline+Next+ tag.

Up to this point, the examples we presented are definable with some
effort in previous work on session types. The following examples
exploit features that make \algst stand out:
\begin{itemize}
\item Traditional session types lack composability in the sense that
  abstraction over behavior is only possible in the tail
  position. Polymorphic context-free session types
  \cite{DBLP:journals/corr/abs-2106-06658} lift this restriction, but
  at the price of a type equivalence
  algorithm that takes doubly exponential time in the
  worst-case. \algst allows abstraction of any part of a protocol
  while providing a linear-time algorithm for type equivalence.
\item The protocol kind is a unique feature of \algst that enables
  abstraction over entire subprotocols and their orientation at any place in a
  protocol. This facility enables us to define a toolbox of generic
  protocol operators, inspired by regular language operations, and
  flip between client and server views using the negation operator.
\end{itemize}

\subsubsection{Generic servers}
\label{sec:generic-servers}

Parametric protocols enable the construction of generic servers that abstract over the
subsidiary protocols. Our first example created a stream of integers,
but what if we wanted a server for \lstinline+Stream Arith+ that  runs the \lstinline+Arith+ protocol ad infinitum?
%
To this end we write a generic server, \lstinline+stream+, that implements
the \lstinline+Stream a+ protocol, for any \lstinline+a+, and invokes a subsidiary server for the
sub-protocol \lstinline+a+. We abstract the type of this
sub-server with a parameterized type alias:
\begin{lstlisting}
type Service a = forall(s:S). ?a.s -> s
\end{lstlisting}
Thus armed, we can write the generic server and use it to run
\lstinline+Stream Arith+.
\begin{lstlisting}
stream : forall(a:P). Service a -> ?Stream a.EndT -> Unit
stream [a] serve c = match c with {
    Next c -> serve [?Stream a.EndT] c |> stream [a] serve }

streamArith : ?Stream Arith.EndT -> Unit
streamArith = stream [Arith] serveArith
\end{lstlisting}
We call the server constructed in this way \emph{passive} as it waits to receive a \lstinline+Next+
tag before running the subsidiary protocol. Dually, we can define an
\emph{active} server. 

\subsubsection{Active servers --- an exercise in duality and negation}
\label{sec:active-servers-an}


An active server offers the
\lstinline+Next+ tag to its client, indicating its readiness to accept a
request. Such a server generalizes the \lstinline+ones+ server.
All we need to do is flip two bits in the type:
\begin{lstlisting}
streamAct : forall(a:P). Service a -> !Stream -a.EndT -> Unit
streamAct [a] svc c =
  select Next [-a, EndT] c |> svc [!Stream -a.EndT] |> streamAct [a] svc
\end{lstlisting}
Why do we have to write  \lstinline+!Stream -a+ in the type?
One might expect the active server to run on a channel
whose type is the dual of \lstinline+?Stream a.EndT+, which is
\lstinline+!Stream a.EndW+.
But this change of direction also affects the
subsidiary protocol in parameter \lstinline+a+ as illustrated by the
typing of \lstinline+select+:
% If \lstinline+c+ has type \lstinline+!Stream Arith.end+, then
\begin{lstlisting}
  c : !Stream Arith.EndT |-
  select Next [Arith, EndT] c : !Arith.!Stream a. EndT
\end{lstlisting}
But this type is not an instance of the type
\lstinline+Service Arith = forall(s:S). ?Arith.s+.
Thanks to polarities, we can adapt to \lstinline+c : !Stream -Arith+ which switches the
direction to an output:
\begin{lstlisting}
  select Next [-Arith, EndT] c : ?Arith.!Stream -Arith. EndT
\end{lstlisting}

This way, we can reuse \lstinline+serveArith+ as follows.
\begin{lstlisting}
streamActArith : !Stream -Arith.EndT -> Unit
streamActArith = streamAct [Arith] serveArith
\end{lstlisting}

The same active server can also serve the infinite stream of \lstinline+ones+:
\begin{lstlisting}
streamActOnes : !Stream Int.EndT -> Unit
streamActOnes = streamAct [-Int] (sendInt 1)
\end{lstlisting}
Interestingly, the typing of \lstinline+streamActOnes+ involves \emph{double negation of
polarities}. Observe that \lstinline+sendInt 1 : Service -Int+, so that the transmitted stream is
in fact \lstinline+Stream -(-Int)+ $\equiv$ \lstinline+Stream Int+.

\subsubsection{A toolbox for generic servers}
\label{sec:toolb-gener-serv}

Inspired by the protocols \lstinline+Stream a+ and \lstinline+ListP a+
(from \cref{sec:introduction}), we can build a library of parameterized 
protocols with generic servers for them. These protocols reify the
standard type constructors, reminiscent of regular language operators.
\begin{lstlisting}
protocol Seq a b    = Seq a b                  -- product
protocol Either a b = Left a | Right b         -- sum
protocol Repeat a   = More a (Repeat a) | Quit -- iteration
\end{lstlisting}
We leave the straightforward implementation of the corresponding passive
generic servers, \lstinline+seq+, \lstinline+either+, and
\lstinline+repeat+, as well as their active variants, to the reader.

These protocol types and their generic servers are sufficient to
construct skeleton servers for all protocols whose flow can be described by regular languages. For example, we
can build the server for \lstinline+Repeat Arith+ by combination of
generic servers with the servers for negation and addition.
\begin{lstlisting}
type Neg    = Seq Int -Int
type Add    = Seq Int (Seq Int -Int)
type Arith  = Either Neg Add

serveNeg : Service Neg
serveNeg [s] c = match c with {
  Seq c -> let (x, c) = receiveInt [!Int.s] c in
           sendInt [s] (0-x) c }

serveAdd : Service Add
serveAdd [s] c = match c with {
  Seq c -> let (x, c) = receiveInt [?Seq Int -Int.s] c in
           match c with {
    Seq c -> let (y, c) = receiveInt [!Int.s] c in
	     sendInt [s] (x+y) c }}

serveArith : Service Arith
serveArith = either [Neg] serveNeg [Add] serveAdd

serveAriths : Service (Repeat Arith)
serveAriths = repeat [Arith] serveArith
\end{lstlisting}

There is some overhead compared to the arithmetic server shown in
\cref{sec:simple-protocols-add}. Previously, a client only had to handle the tags
\lstinline+Neg+ and \lstinline+Add+ besides the \lstinline+Repeat+
protocol. Here, \lstinline+Neg+ and \lstinline+Add+ are replaced by
\lstinline+Left+ and \lstinline+Right+ and the client has to handle
the \lstinline+Seq+ tags on top. Thus the protocol constructed from
the generic parts has some overheads, as we discuss in
\cref{sec:are-generic-servers}. We provide further examples of 
generic servers in \cref{sec:toolb-gener-serv-1}.


% \subsubsection{Transmission of parametric data}
% \label{sec:transm-param-data}


% Using parameterized protocols, we can define polymorphic operations for sending and receiving
% parametric datatypes. Suppose we want to send lists of arbitrary type \lstinline+a+, as long as someone can
% supply a way of sending a value of type \lstinline+a+.
% \begin{lstlisting}
% data List (a:T) = Nil | Cons a (List a)
% type Send (a:T) = a -> forall(s:S). !a.s -> s

% sendList : forall(a:T). Send a -> Send (List a)
% sendList [a] sendA xs [s] c = case xs of {
%     Nil       -> select Nil [a,s] c,
%     Cons x xs -> select Cons [a,s] c
%               |> sendA [!List a.s] x
% 	      |> sendList [a] sendA [s] xs }
% \end{lstlisting}
% To this end, one would provide implementations of \lstinline+Send a+
% for suitable primitive types \lstinline+a+ and modularly build further
% implementations.
% % \begin{lstlisting}
% % sendInt : Send Int
% % sendInt x [s] = send [Int, s] x
% % \end{lstlisting}
% \begin{lstlisting}
% sendListInt : Send (List Int)
% sendListInt = sendList [Int] sendInt

% sendListListInt : Send (List (List Int))
% sendListListInt = sendList [List Int] sendListInt
% \end{lstlisting}
% %Receiving can be handled similarly.


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
