\section{Modelling Recursive Session Types via Algebraic Protocols}

This section describes how recursive session types \`a la
\citet{DBLP:journals/jfp/GayV10} can be embedded in algebraic protocols. To
differentiate the two different type languages we use ``algebraic types'' to
refer to session and protocol types as defined by
\citet{DBLP:journals/pacmpl/MordidoS0V23}, ``recursive types'' refer to
classical, end-recursive session types. Both notions also include the
corresponding functional types.


\subsection{A graph representation for recursive types}

Recursive session types are represented as a deterministic labelled transition
system (LTS) $\stlts$ where $V$ is a set of nodes or states, $L$ is the set of
transition labels containing at least $\Lone,\Ltwo\in L$, and ${\sttrans{}}
\subseteq V\times L\times V$ the transition relation. We may write
$v\sttrans\ell v'$ for $(v,\ell,v')\in{\sttrans{}}$, and $v\notrans$ if there
are no $\ell,v'$ such that $v\sttrans\ell v'$. The labelling function
$\lab:V\to\labset$ assigns a label to every state. The set of state labels must
comprise at least the so called set of session type labels $\labset \supset
\labsetS = \{\LSend,\LRecv,\LChoice,\LBranch,\LEnd\}$ and the function arrow
$\LFunc\in\labset$. The set of base type labels $\labsetB$ is defined as the
set of remaining labels $\labsetB = \labset \setminus \labsetS \setminus
\{\LFunc\}$. Typical candidates for the set of base types are, for example
$\LUnit$ or $\LInt$. Every node $v\in V$ must satisfy the conditions below.
%
\begin{enumerate}
  %
  \item \label{tylts:arity0} If $\lab(v) = \LEnd$ or $\lab(v) \in \labsetB$
    it must hold that $v\notrans$.
  %
  \item \label{tylts:arity2} If $\lab(v) \in \{\LSend,\LRecv,\LFunc\}$ there
    must be $v_1,v_2\in V$ with $v \sttrans\Lone v_1$ and $v \sttrans\Ltwo
    v_2$, and for all $v \sttrans\ell v'$ it must hold that
    $(\ell,v')\in\{(\Lone,v_1),(\Ltwo,v_2)\}$.
  %
  \item If $\lab(v)\in\{\LChoice,\LBranch\}$ there must be at least one
    transition $v\sttrans\ell v'$, and for all transitions $v \sttrans{\ell'}
    v''$ it must hold that $\ell'\not\in\{1,2\}$.
  %
  \item If $\lab(v)\in\labsetS$, for all $v\sttrans\ell v'$ with $\ell \ne
    \Lone$ it must hold that $\lab(v')\in\labsetS$.
  %
  \item \label{tylts:cycles} If there is a cycle $v \sttrans{\ell} v_1
    \sttrans{\ell_1} \cdots v_n \sttrans{\ell_n} v_1$ it must hold that
    $\lab(v)\in\labsetS$ and $\ell\ne\Lone$.
  %
  \item If $v\sttrans\ell v'$ and $v\sttrans\ell v''$ it must hold that
    $v'=v''$.
  %
\end{enumerate}
%
Requirement (\ref{tylts:cycles}) restricts the kind of cycles that may occur.
The only cycles that are allowed must consists solely of session type nodes and
the transition labels must not be $\Lone$. The restriction serves two goals:
only sessions types are allowed to be recursive and we avoid the dualization
paradox where recursive session types use recursion variables in the payload of
messages~\citep{BernardiH16,DBLP:journals/corr/abs-2004-01322}.

Given an LTS $\stlts$ the function $\stty{{\,\cdot\,}}$ produces a type with
the grammar below. In particular, if $\lab(v)\in\labsetS$ for some $v\in V$,
the result of $\stty{v}$ is a type according to the grammar $S$. To
differentiate the translation functions for two transition systems $A$ and $B$
we may write $\stty{{\,\cdot\,}}^A$ and $\stty{{\,\cdot\,}}^B$ respectively.
%
\begin{align*}
  %
  T &\grmeq \dots \\
  %
  S &\grmeq \dots
  %
\end{align*}
%
% $\eta$ is a partial function from nodes to recursion variables. We write
% $\eta(v)\downarrow$ if $\eta$ is defined at $v$, and $\eta(v)\uparrow$ if
% not. $\emptyset$ denotes the function that is not defined at any point and
% $\eta[v\mapsto\alpha]$ the partial function defined as $\alpha$ at point $v$
% and defined according to $\eta$ at any other point.

\[
  \stty{v} = \begin{cases*}
    \recc v & if
      $v \in \eta$
    \\
    \SRec[v] \SBranchChoice\star*{{\ell: \stty[\eta\cup\{v\}]{v'}}}_{\recc{\ell\in I}} & if
      %$\eta(v)\uparrow$,
      $v \not\in \eta$,
      $\star = \lab(v) \in \{\LBranch,\LChoice\}$,
      %$\alpha$ fresh,
      %for all $\ell,v'$ with $v \sttrans\ell v'$
      $I = \{ \ell\in L \mid \exists v'\!.\, v \sttrans\ell v' \}$
    \\
    \SRec[v] \SSendRecv\sharp{\stty[\emptyset]{v_1}} \stty[\eta\cup\{v\}]{v_2} & if
      %$\eta(v)\uparrow$,
      $v \not\in \eta$,
      $\sharp = \lab(v) \in \{\LSend,\LRecv\}$,
      %$\alpha$ fresh,
      $v \sttrans{\Lone} v_1, v \sttrans{\Ltwo} v_2$
    \\
    \recc{\stty[\emptyset]{v_1} \to \stty[\emptyset]{v_2}} & if
      $\lab(v) = \LFunc$,
      $v \sttrans{\Lone} v_1, v \sttrans{\Ltwo} v_2$
    \\
    \SEnd & if $\lab(v) = \LEnd$ \\
    \lab(v) & if $\lab(v) \in \labsetB$ \\
  \end{cases*}
\]
%
In the last clause of we assume that every base type label corresponds to a
valid base type. This assumption is reasonable for choices like $\LUnit$ or
$\LInt$.

\begin{lemma}
  %
  The types produced by $\stty{\,\cdot\,}$ are well-formed, that is every type
  is contractive and formed according to the grammar $T$. An argument with a
  label in $\labsetS$ forms a type according to the grammar $S$.
 %
\end{lemma}
%
\begin{proof}
  %
  By inspection of $\stty{\,\cdot\,}$ every produced type is contractive.
  \js{incomplete}
  %
\end{proof}
%
\begin{lemma}
  %
  The types produced by $\stty{\,\cdot\,}$ contain recursion variables only in
  tail recursive positions of session types.
  %
\end{lemma}
%
\begin{proof}
  %
  \js{TODO}
  %
\end{proof}


\subsection{Duality}

We need a way to identify nodes in the LTS that represent dual types. To this
end we define the notion of duality relations.

\begin{definition}
  Given an LTS $\stlts$ a relation ${\DualRel} \subseteq V\times V$ is a
  \emph{duality relation} if for any two nodes $u,v \in V$ with $u\DualRel v$
  it holds that

  \begin{enumerate}
    \item\label{dual-rel/labels} $\{\lab(u),\lab(v)\} \in \{ \{\LSend,\LRecv\},
      \{\LChoice,\LBranch\} \}$
    \item\label{dual-rel/u2v} $u \sttrans\ell u'$ for $\ell\in
      L\setminus\{\Lone\}$ implies $v \sttrans\ell v'$ such that $u' \DualRel
      v'$
    \item\label{dual-rel/v2u} $v \sttrans\ell v'$ for $\ell\in
      L\setminus\{\Lone\}$ implies $u \sttrans\ell u'$ such that $u' \DualRel
      v'$
    \item\label{dual-rel/one} $u\sttrans\Lone w$ if and only if $v\sttrans\Lone
      w$ \js{this formulation requires a form of uniqueness for nodes}
  \end{enumerate}
\end{definition}

\begin{lemma}
  Any duality relation ${\DualRel}$ is symmetric.
\end{lemma}
\begin{proof}
  If requirements (\ref{dual-rel/labels}) and
  (\ref{dual-rel/one}) are satisfied they will also be trivially satisfied in
  the symmetric case. Condition (\ref{dual-rel/u2v}) guarantees the condition
  (\ref{dual-rel/v2u}) is satisfied for the symmetric case and vice versa.
\end{proof}

\begin{definition}
  Given an LTS $\stlts$, two nodes session type nodes $u,v\in V$ are called
  \emph{dual} if there exists a duality relation ${\DualRel}$ such that
  $u\DualRel v$.
\end{definition}

\begin{lemma}
  Given two dual nodes $u,v$ then $\SDual{\stty{u}} = \stty{v}$.
\end{lemma}
\begin{proof}
  \js{TODO}
\end{proof}

\begin{definition}
  An LTS is called \emph{uniquely labelled} if for every pair of nodes $u,v$
  ($u\ne v$) the existence of a transition label $\ell\not\in \{\Lone,\Ltwo\}$
  such that $u \sttrans\ell u'$ and $v\sttrans\ell v'$ implies that $u$ and $v$
  are dual nodes.
\end{definition}

\subsection{Subtyping}

We define a subtyping relation on the LTS representation and prove it sound
with respect to the classical subtyping relation for session types
\citep{DBLP:journals/acta/GayH05,DBLP:journals/jfp/GayV10}.

Subtyping is defined on the transition relation of the LTS. Given two recursive
processes $T = \stlts[T]$ and $U = \stlts[U]$ over the same set of nodes and
labels, and with the same labelling function, we define $T \SubLTS U$ by
$\SubNode{v_0}(\sttrans{}_{T}, \sttrans{}_{U})$, where $v_0$ is the initial
node. The binary relation $\SubNode{v}(\sttrans{~}_{A}, \sttrans{~}_{B})$ with
parameter $v \in V$ is defined via a case split on the label of $v$.
%
\begin{itemize}

  \item \textbf{Case $\lab(v) = \LBranch$:} For every transition $v \sttrans{\ell}_{A} v'$,
        there must be a matching transition $v \sttrans{\ell}_{B} v'$ such that
        $\SubNode{v'}(\sttrans{}_{A}, \sttrans{}_{B})$.

  \item \textbf{Case $\lab(v) = \LChoice$:} For every transition $v \sttrans{\ell}_{B} v'$,
        there must be a matching transition $v \sttrans{\ell}_{A} v'$ such that
        $\SubNode{v'}(\sttrans{}_{A}, \sttrans{}_{B})$.

  \item \textbf{Case $\lab(v) = \LSend$:} There must be matching transitions
        $v \sttrans{\Lone}_{A} v'_1$, $v \sttrans{\Lone}_{B} v'_1$ and
        $v \sttrans{\Ltwo}_{A} v'_2$, $v \sttrans{\Ltwo}_{B} v'_2$ such that
        $\SubNode{v'_1}(\sttrans{}_{B}, \sttrans{}_{A})$ and
        $\SubNode{v'_2}(\sttrans{}_{A}, \sttrans{}_{B})$.

  \item \textbf{Case $\lab(v) = \LRecv$:} There must be matching transitions
        $v \sttrans{\Lone}_{A} v'_1$, $v \sttrans{\Lone}_{B} v'_1$ and
        $v \sttrans{\Ltwo}_{A} v'_2$, $v \sttrans{\Ltwo}_{B} v'_2$ such that
        $\SubNode{v'_1}(\sttrans{}_{A}, \sttrans{}_{B})$ and
        $\SubNode{v'_2}(\sttrans{}_{A}, \sttrans{}_{B})$.

  \item \textbf{Case $\lab(v) = \LFunc$:} There must be matching transitions
        $v \sttrans{\Lone}_{A} v'_1$, $v \sttrans{\Lone}_{B} {v'_1}$ and
        $v \sttrans{\Ltwo}_{A} v'_2$, $v \sttrans{\Ltwo}_{B} {v'_2}$ such that
        $\SubNode{v'_1}(\sttrans{}_{B}, \sttrans{}_{A})$ and
        $\SubNode{v'_2}(\sttrans{}_{A}, \sttrans{}_{B})$.

  \item \textbf{Case $\lab(v) = \LEnd$:} The relation holds trivially.
\end{itemize}

For $\LChoice$, the direction of the implication requires the subtype to
provide at least all the transitions offered by the supertype, meaning the
smaller type can support more labels. Conversely, for $\LBranch$, the subtype
is restricted to taking only transitions permitted by the supertype, meaning
the smaller type must have fewer or an equal amount of labels. Furthermore, for
$\LSend$ and $\LRecv$, the continuation of the first argument (label $\Lone$)
is contravariant and covariant, respectively. Similarly, the argument (label
$\Lone$) for $\LFunc$ is contravariant. Finally, $\LEnd$ holds trivially
because it admits no transitions.

\begin{lemma}[Soundness of subtyping]
  %
  Given two typing LTS $A,B$ defined over the same set of nodes $V$ and with
  the same labelling function $\lab\!$, if
  $\SubNode{v}(\sttrans{}_{A},\sttrans{}_{B})$ for $v\in V$ then $\stty{v}^A
  \SubST \stty{v}^B$.
  %
\end{lemma}
%
\begin{proof}
  \js{TODO}
\end{proof}


\subsection{Recursive types as protocols}

This section describes how a recursive session type is translated to a valid
protocol type. The types are assumed to be available in their LTS
representation described previously. A node in the LTS labelled by an element
of $\labsetS$ is translated to a protocol type. Other nodes are translated to
types of kind $\kindt$ by embedding protocol types $\typec P$ as session types
$\typec{!P.\TEnd}$.

\begin{lstlisting}[
  float,
  caption={Auxiliary protocols used in the translation to algebraic protocols.},
  label=lst:translation-protocols,
  emph={Eps,Seq,Rep},
  frame=single
]
protocol Eps      = Eps
protocol Seq p q  = Seq p q
protocol Rep p    = Rep p (Rep p)
\end{lstlisting}

The translation makes use of the three singleton protocols shown in
\cref{lst:translation-protocols}. Protocol $\PEps$ is operationally a no-op,
$\TPApp{\TPApp{\PSeq}{p}}{q}$ sequences two protocols $\typec p$ and $\typec
q$, and $\TPApp\PRep p$ repeats protocol $\typec p$ ad infinitum. They are used
to encode recursive session types that do not contain branch or choice
constructors. For example, the session type the session type
$S_{\rightleftharpoons} = \recc{\SRec\SSend\TInt\SRecv\TInt\alpha}$ will be
encoded as
$\typec{!\TPApp\PRep{(\TPApp{\TPApp\PSeq{\ppos\TInt}}{(\TPApp{\TPApp\PSeq{\pneg\TInt}}{\PEps})})}.\TEnd}$.

All other session types are encoded as a potentially empty sequence of sends
and receives, followed optionally by a single send or receive of a global
$\PDispatch$ protocol synthesized by the translation, and terminated by
$\TEnd$. In this case, any recursion is ``hidden'' inside the $\PDispatch$
protocol. The translation of whole programs only synthesizes one such protocol.
Protocol subtyping is used to restrict the set of constructors that are
actually valid. For example the session type $\SSend\SInt\SRec\SChoice{
\textsf{Add}:{\SSend\SInt\alpha}, \textsf{Result}:{\SRecv\SInt\SEnd} }$ is
encoded as $\typec{!\TInt.!\PDispatch[\textsf{Add},\textsf{Result}].\TEnd}$
where $\PDispatch$ has at least the following constructors:
%
\begin{lstlisting}[emph={Dispatch,Int,Eps},emph={[2]Add,Result},emphstyle={[2]\color{ACMDarkBlue}}]
protocol Dispatch = ...
                  | Add Int Dispatch[Add, Result]
                  | Result -Int Eps
\end{lstlisting}

For a sequence of of nodes $s = v_1,v_2,\dotsc,v_n$ the $i$-th rotation ($1\leq
i\leq n$), written as $\seqrotate i s$, denotes the sequence
$v_i,v_{i+1},\dotsc,v_n,v_1,\dotsc,v_{i-1}$. Given a total order on the nodes
${\VLeq} \subseteq V \times V$ the \emph{minimal rotation} is an $1\leq i\leq
n$ such that for any $1\leq j\leq n$ it holds that $\seqrotate i s \VLeq
\seqrotate j s$ by lexicographically extending $\VLeq$ to sequences.

The function $\ToAlgPM{}$ maps a subset of node labels into type polarities. It
is lifted to nodes such that $\ToAlgPM{v}=\ToAlgPM{\lab(v)}$ under the
condition that $v$ is suitably labelled.
%
\begin{align*}
  \ToAlgPM{} &\centeronequals{$:$} \{\LSend,\LRecv,\LChoice,\LBranch\} \to \{\ppos{},\pneg{}\} \\
  \ToAlgPM{} & = \left\{
    \!\!\begin{array}{l@{}l@{{}\mapsto{}}l}
      \LSend,&\LChoice & \ppos{} \\
      \LRecv,{}&\LBranch & \pneg{} \\
    \end{array}
  \right.
\end{align*}

\begin{figure}
  \newcommand\ToAlgTyProtoClause{\typec{! \ppm{} \ToAlgProto{v,\{v'\mapsto 0\mid v'\in V\}} .\TEnd}}
  \small
  \begin{align*}
    \ToAlgTy{v} &= \begin{cases*}
      \lab(v) & if
        $\lab(v) \in \labsetB$
      \\
      \TFun[] {\ToAlgTy{v_1}} {\ToAlgTy{v_2}} & if
        $\lab(v) = \LFunc$,
        $v \sttrans{\Lone} v_1, v \sttrans{\Ltwo} v_2$,
      \\
      \TEnd & if
        $\lab(v) = \LEnd$
      \\
      \ToAlgTyProtoClause
            & if $\lab(v)\in\labsetS\setminus\{\LEnd\}$, ${\ppm{}} = \ToAlgPM{v}$
      \\
    \end{cases*}
    \\[\abovedisplayskip]
    \ToAlgProto{v,\ToAlgSeen} &= \begin{cases*}
      \ToAlgStep{v,\ToAlgUpdate v 1} & if $\ToAlgSeen(v) = 0$ \\
      \PRep\ \ToAlgStep{v,\ToAlgUpdate v 2} & if $\ToAlgSeen(v) = 1$ \\
      \mathrlap{\PEps}\hphantom{\ToAlgTyProtoClause} & if $\ToAlgSeen(v) = 2$
    \end{cases*}
    \\[\abovedisplayskip]
    \ToAlgStep{v,\ToAlgSeen} &= \begin{cases*}
      \mathrlap{\PEps}\hphantom{\ToAlgTyProtoClause}
        & if $\lab(v) = \LEnd$ \\
      \PSeq\ \ppm{}\ToAlgPM{v}\ToAlgTy{v_1}\ \ToAlgProto{v_2,\ToAlgSeen}
        & if $\lab(v) \in \{\LSend,\LRecv\}$,
           $v \sttrans{\Lone} v_1$,
           $v \sttrans{\Ltwo} v_2$
        \\
      \ppm{}\ToAlgPM{v}\PDispatch[I]
        & if $\lab(v) \in \{\LChoice,\LBranch\}$,
          $ I = \{\ell\in L \mid \exists v'\!.\,v \sttrans\ell v'\}$
    \end{cases*}
  \end{align*}
  \caption{Translation functions}
  \label{fig:translation-functions}
\end{figure}

Nodes in the LTS are translated to algebraic types through the three mutually
recursive functions $\ToAlgTy{}$, $\ToAlgProto{}$, $\ToAlgStep{}$ defined in
\cref{fig:translation-functions}. $\ToAlgTy{}$ is the main entry point. The
latter two functions translate session types to algebraic protocols. A call to
$\ToAlgTy{}$ always terminates because the only cycles in the LTS must be
through session types nodes. The call to $\ToAlgProto{}$ always terminates
because it terminates at the latest when a node is encountered for the third
time.

The protocol $\PDispatch$ is defined as
%
\[
  \protocolk\ \PDispatch = \left\{
    \termc{\constr_\ell}\ \cramped{\ToAlgProto[\ToAlgPM{v}]{v'}} \;\middle|\;
    v,v' \in V,\ \ell\in L\setminus\{\Lone,\Ltwo\},\ %
    v\sttrans\ell v'
  \right\}
\]

\begin{lemma}[Well-formedness of generated types]\
  \begin{enumerate}
    \item For all $v\in V$ it is derivable that $\typeAgainst {\ToAlgTy{v}}
      \kindt$. In particular, $\lab(v) \in \labsetS$ implies $\typeAgainst
      {\ToAlgTy{v}} \kinds$ is derivable.
    \item For all $v\in V$ where $\lab(v)\in\labsetS$ and all\/ $\ToAlgSeen : V \to
      \{0,1,2\}$ it is derivable that $\typeAgainst {\ToAlgProto{v,\ToAlgSeen}}
      \kindp$ and $\typeAgainst {\ToAlgStep{v,\ToAlgSeen}} \kindp$.
  \end{enumerate}
  In either case it is assumed that $\Delta$ contains at least the protocol
  definitions for $\PEps$, $\PSeq$, $\PRep$, $\PDispatch$.
\end{lemma}

\begin{lemma}
  Given two LTS $A=\stlts[A]$ and $B=\stlts[B]$ defined over the same set of
  nodes, transition labels, and the same labelling function, for all $v\in V$
  it holds that $\SubNode v(\sttrans{}_A,\sttrans{}_B)$ if and only if\/
  $\ToAlgTy[A]{v} \subt \ToAlgTy[B]{v}$.
\end{lemma}

\begin{lemma}
  For all pairs of session type nodes $u,v$ it holds that $u$ and $v$ are dual
  nodes if and only if $\Normal[-]{\ToAlgTy{u}}=\Normal[+]{\ToAlgTy{v}}$.
\end{lemma}


\subsection{Expressions and processes}

After having specified how the type language of end-recursive session types can
be embedded in algebraic protocols we now tackle the translation of expressions
and processes. The translation mainly needs to insert additional $\selectk$ and
$\matchk$ statements to handle the auxiliary protocols introduced by the type
translation.

Due to the way session types are translated a $\appe \selectk \CSeq$ or
corresponding $\matchk$ must be inserted in front of each $\sendk$ and
$\receivek$. Already existing $\selectk$ and $\matchk$ operations correspond
directly to their counterparts on the $\PDispatch$ protocol. Since $\SEnd$ is
translated to itself but $\SSend\SInt\SEnd$ is translated to
$\typec{!\TPApp{\TPApp\PSeq\TInt}{\PEps}.\TEnd}$ which leaves
$\typec{!\PEps.\TEnd}$ after the communication step $\selectk$ and $\matchk$
statements must be introduced to deal with $\PEps$ protocols. Similar handling
is necessary for non-branching recursive session types like
$\SRec\SSend\SInt\recc\alpha$.
